### 2022-05-18 paper/paper.tex
methodology: add methodology

@@

@@

\chapter{Advanced Review of Related Literature}

@@

The experimental setup and configuration will be composed of three elements: the
camera, the keyboard, and the environment. Figure~\ref{fig:metho-setup}
shows the sketch of the complete experimental setup.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.4\textwidth]{setup.png}
	\caption{Sketch of the experimental setup}
	\label{fig:metho-setup}
	\centering
\end{figure}

\subsection{Camera}
The camera setup will use a single monocular camera positioned in a top-down
view. The camera will capture the entirety of the keyboard and the movement of
the ten fingers. To do so, it will be mounted on top of the monitor and the
camera will point down towards the table. Figure~\ref{fig:metho-setup-keeb}
illustrates what the camera will capture once placed in its correct position.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{actual-keeb.png}
	\caption{Keyboard, camera angle, and placement to be used for testing.}
	\label{fig:metho-setup-keeb}
\end{figure}

The specific camera to be used will be a Logitech C920.
Figure~\ref{fig:metho-setup-cam} is a picture of this camera. It is a 3 mega pixel
webcam that is capable of capturing color video in 1080p/30fps and 720p/30fps
with a diagonal field of view of 78\degree. This camera has a universal mounting
clip that will allow the camera to be correctly positioned within the
experimental setup \parencite{logitech}.

@@

The keyboard will be a 60\% keyboard as shown in
Figure~\ref{fig:metho-setup-keeb}. This will limit the necessary mapping for the
algorithm to the alphanumeric portion of the keyboard. This keyboard choice will
also lessen the area that the camera will need to capture, as this keyboard type
is considerably smaller compared to a full-size keyboard. The keyboard will have
keycaps and a case in a light color that will contrast the dark surface that it
will be placed on. This is to improve initial keyboard detection. In addition,
the keyboard layout will also be ANSI, due to the availability and widespread
adoption of the layout in the Philippines.

\subsection{Environment}
The setup would be placed in would be a well lit environment with a light source
beside the camera. This light source will evenly light the keyboard and the
fingers used for typing. The light source to be used will be a common LED desk
lamp rated at 9 Watts with 700 lumens. This light will be white with a color
temperature of 6500k.

In addition, the surface where the keyboard is to be placed on will be solid
black without any variation of color. This is to improve initial keyboard
detection.

@@

\subsection{Computer Vision based Keyboard Detection, and Mapping}
\label{section:metho-algo-keyboard}

A computer vision algorithm will be created as a starting point for detecting,
and mapping the keyboard within a video. The flowchart of the algorithm is shown
in Figure~\ref{fig:metho-algo-key-flow}.

@@

		\node[draw,
			below=of gray,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (sobel) {Edge Detection using Sobel Filter};

		\node[draw,
			below=of sobel,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (contours) {Find Contours (Simple)};

		\node[draw,
			below=of contours,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (map) {Overlay Virtual Keyboard Map using Contour Points};

		\node[draw,
			trapezium,
			trapezium left angle = 65,
			trapezium right angle = 115,
			trapezium stretches,
			below=of map,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (output) {Output Frame};

@@

		\draw[-latex] (start) edge (input)
		(input) edge (gray)
		(gray) edge (sobel)
		(sobel) edge (contours)
		(contours) edge (map)
		(map) edge (output)
		(output) edge (end);

\subsubsection{Convert to Grayscale}
The frame will be converted to a 256 level grayscale image. This step will be
performed because future steps of the algorithm will not require color values to
work. In addition, the performance of the following steps will be also improved
as the number of dimensions to be analyzed is reduced.

\subsubsection{Edge Detection using Sobel Filter}
A Sobel Filter is an edge detection filter that uses a 3$\times$3 kernel that is
convolved twice. Once horizontally, and another vertically to produce a
grayscale image of the outlines within the frame. The kernels used by the Sobel
Filter \parencite{sobel2014} is shown in Figure~\ref{fig:metho-algo-key-sobel}.

\begin{figure}[H]
	\centering
	$\begin{bmatrix}
			+1 & 0 & -1 \\
			+2 & 0 & -2 \\
			+1 & 0 & -1
		\end{bmatrix}$
	and
	$\begin{bmatrix}
			+1 & +2 & +1 \\
			0  & 0  & 0  \\
			-1 & -2 & -1
		\end{bmatrix}$
	\caption{Sobel Operator Kernels. Reproduced from \fullcite{sobel2014}}
	\label{fig:metho-algo-key-sobel}
\end{figure}

The expected result of this step is a black and white image that highlights the
edges of the keyboard in white.

\subsubsection{Find Contours (Simple)}
\label{section:metho-algo-key-contours}
Contours are curves joining all continuous points that have the same color or
intensity \parencite{opencv-contours}. The OpenCV function \texttt{findContours}
will be used to find the contours of the outlines of the object found using the
previous step. This function accepts a contour approximation method as one of
its arguments. Two methods are provided by OpenCV:\@\texttt{CHAIN\_APPROX\_NONE}
and \texttt{CHAIN\_APPROX\_SIMPLE}. The former returns all contours in the
shape, while the latter removes redundant points and returns the least amount of
points that describes the shape. The algorithm will use the latter, as only the
extreme edges of the keyboard needs to be detected. This OpenCV function
implements the algorithm of~\cite{contours} in their paper
\citetitle{contours}.

The expected result of this algorithm is four contour points which are
positioned on the four edges of the keyboard.

\subsubsection{Overlay Virtual Keyboard Map using Contour Points}
The virtual keyboard map is a rectangular image that contains individual
\ac{roi} for each key in a 60\% \ac{ansi} keyboard. Each key has a corresponding
color assigned to it and this color fills the region where this key is located
at. Figure~\ref{fig:metho-algo-key-map} is the virtual keyboard map. Appendix
Chapter~\ref{appendix:key-map} is a table that shows the hex color code and key
mapping present in the image map.

@@

The virtual map will then be stretched over the object, with the four contour
points as the four edges of the virtual keyboard map. The image generated will
then be returned as the final output of the graph.
Figure~\ref{fig:metho-algo-key-overlay} illustrates this image, and
Figure~\ref{fig:metho-algo-key-half} shows how accurate the image map is when
overlaid over the image.

\begin{figure}[H]
	\centering
	\includegraphics[width=1\textwidth]{image-map-final.png}
	\caption{Expected output of the algorithm}
	\label{fig:metho-algo-key-overlay}
	\centering
\end{figure}

\hfill

\begin{figure}[H]
	\centering
	\includegraphics[width=1\textwidth]{image-map-half.png}
	\caption{Overlaid image map shown at 60\% opacity}
	\label{fig:metho-algo-key-half}
	\centering
\end{figure}

\subsection{Computer Vision based Finger Detection, and Tracking}
\label{section:metho-algo-finger}
A computer vision algorithm will be chosen as a starting point for detecting,
and tracking fingers within a video. The specific algorithm used for finger
detection and tracking will be the solution offered by MediaPipe dubbed
MediaPipe Hands. This algorithm is composed of two ML models working in
conjunction to be able to detect the different parts of the hands and track them
accurately.

@@

The two previously chosen algorithms will be combined to accomplish finger-key
identification and mapping. Finger-key identification refers to identifying
which finger is used to press which key. As an example, the key Q was pressed
with the pinky finger of the left hand.

The integration of the algorithm is a two-step process. The first step is to get
the mapping of the keyboard using the algorithm shown in
Section~\ref{section:metho-algo-keyboard}.

The second step runs whenever a key press has been detected. This step uses the
mapping of the keyboard in conjunction with the finger tracking algorithm
selected in Section~\ref{section:metho-algo-finger}. The flowchart of the
algorithm is shown in Figure~\ref{fig:metho-algo-integration}

@@

		\node[draw,
			trapezium,
			trapezium left angle = 65,
			trapezium right angle = 115,
			trapezium stretches,
			below = 2cm of if-packet,
			minimum width = 3.5cm,
			minimum height = 1cm,
			align = center,
			inner sep = 3mm
		] (map) {Virtual Keyboard Map \\ Finger Tracking Data};

		\node[draw,
			below = of map,
			minimum width = 3.5cm,
			minimum height = 1cm,
			inner sep = 3mm
		] (find-roi) {Find \ac{roi} of the pressed key in Virtual Keyboard Map};

		\node[draw,
			below=of find-roi,
			minimum width = 3.5cm,
			minimum height = 1cm,
			inner sep = 3mm
		] (find-finger) {Find which fingertip is on that \ac{roi}};

@@

		\draw[-latex] (start) edge (if-packet)
		(map) edge (find-roi)
		(find-roi) edge (find-finger)
		(find-finger) edge (return)
		(return) edge (end);

@@

\subsubsection{Find \ac{roi} of the pressed key in Virtual Keyboard Map}
Since there is a known color-key mapping based on the Virtual Keyboard Map, the
algorithm will find the correct color to search for based on the received key
press packet. This color is isolated from the Virtual Keyboard Map, and the
\ac{roi} and its coordinates are obtained. The edges of this singular \ac{roi}
will then be obtained by finding its contours. The specific implementation of
this is explained in Section~\ref{section:metho-algo-key-contours}.

The coordinate system will have its origin at the top left, starting at 0, 0.
Going to the right increases the value of the x-axis, and going to the bottom
increases the value of the y-axis. This will result in a coordinate system that
only has positive values with each value corresponding to a pixel.

The expected output of this is four sets of pairs. Each pair corresponds to an
edge of the single \ac{roi}.

\subsubsection{Find which fingertip is on that \ac{roi}}
The Finger Tracking Data will contain the pixel positions of each landmark of
the hand. For this step, the algorithm will find the landmark which is
positioned within the single \ac{roi} obtained from the previous step.
This will be done using a series of checks.

Each landmark's coordinates will be compared to the coordinates of the edges of
the \ac{roi}. A landmark will be determined as positioned within the \ac{roi} if all the
following conditions are true:
\begin{enumerate}
	\item In the X axis, the landmark's coordinates is greater than one or both of
	      the two coordinates found of the left side of the \ac{roi}
	\item In the X axis, the landmark's coordinates is less than one or both of
	      the two coordinates found of the right side of the \ac{roi}
	\item In the Y axis, the landmark's coordinates is greater than one or both of
	      the two coordinates found of the top side of the \ac{roi}
	\item In the Y axis, the landmark's coordinates is less than one or both of
	      the two coordinates found of the bottom side of the \ac{roi}
\end{enumerate}

These conditions maximize the total area of the \ac{roi} and is not strict about
exact accuracy. In essence, these conditions will create a perfectly rectangular
box that contains the quadrilateral formed by the four points from the previous
step.

\subsubsection{Return which fingertip}
The landmark will then be used to determine which specific finger corresponds to
the key press. Figure~\ref{fig:metho-algo-integration-landmarks} shows each
possible landmark that may be returned from the previous step. This step will
then return the name of the landmark, up until the first underscore. As an
example, if the landmark found is \texttt{MIDDLE\_FINGER\_TIP}, this step will
return \texttt{MIDDLE} denoting that the middle finger is the finger that
corresponds with the key press. In addition, the specific hand will also be
returned as one of two strings, \texttt{Left} and \texttt{Right}, since
this information is also bundled together with the landmarks.

@@

\subsection{Implementation}
\label{section:metho-algo-implementation}
The previously discussed algorithms will be implemented using Python. The
Keyboard Detection, and Mapping Algorithm in
Section~\ref{section:metho-algo-keyboard} will be implemented as a separate
submodule in Python using OpenCV.\@The Finger Detection, and Tracking Solution
in Section~\ref{section:metho-algo-finger} will be consumed using the prebuilt
Python package offered by the MediaPipe team. Finally, the integration of the
two will be done as another separate submodule in Python.

\section{Trainer}

\subsection{Typing Test Sequences}
Typing test sequences are strings that will be used in testing the user in their
ability to type. The test sequences that will be used in the trainer will come
from text found in the public domain obtained from Project Gutenberg and the
Internet Archive. Sentences will be isolated from these text and used as test
sequences if they fit the criteria.

The criteria for choosing test sequences is as follows: (1) $\approx80\%$ of the
characters in the keyboard is present in a test sequence. (2) The number of
words in a test sequence do not exceed 25. (3) Numbers and punctuations should
be present in at least $\approx30\%$ of the total test sequences.

There will be a total of 10 test sequences that will be gathered. An example
test sequence is ``What of it, if some old hunks of a sea-captain orders me to
get a broom and sweep down the decks?'' from Moby Dick by~\cite{moby-dick}. The
9 other test sequences can be found in Appendix
Chapter~\ref{appendix:test-sequences}.

\subsection{UX/UI design and Front-end development for the touch typing trainer}
\subsubsection{Design}
This design will be based on other type tests such as Monkeytype
\parencite{bartnik2021}, and Keybr \parencite{keybr}, with additional components
for showing real-time finger-key identification and mapping.
Figure~\ref{fig:ui-test} and~\ref{fig:ui-stat} illustrates the two main pages of
the design. Additional screens, such as historical statistics, in-platform
tutorial, and a written guide are shown in the appendix.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{ui-test.png}
	\caption{Test screen of the front-end design}
	\label{fig:ui-test}
	\centering
\end{figure}

Figure~\ref{fig:ui-test} has four main components. From the top, the first
component is the navigation bar. This directs the user to the different pages
within the website. The second component in the upper left displays the number
of remaining words to be typed. This is 16 in the example. Below that is the
test sequence. This shows the user what characters to type, and highlights
mistakes in two distinct colors. Red if the character pressed was incorrect and
purple if the character typed was correct, but the finger used to type it is
wrong. The last main component in this screen is the keyboard that illustrates
the correct finger positioning, with each finger corresponding to a different
color.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{ui-stats.png}
	\caption{Statistics screen of the front-end design}
	\label{fig:ui-stat}
	\centering
\end{figure}

>>>>

\usepackage{caption}

@@

\listoftables
\newpage

@@

\chapter{Review of Related Literature}

@@

The experimental setup and configuration was composed of three elements: the
camera, the keyboard, and the environment. Figure~\ref{fig:metho-setup} is an
image of the experimental setup

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{full-setup.jpg}
	\caption{The experimental setup}
	\label{fig:metho-setup}
	\centering
\end{figure}

\subsection{Camera}
The setup used a single monocular camera positioned in a top-down view. The
camera captured the entirety of the keyboard and the movement of the ten
fingers. To do so, it was mounted on top of the monitor with the camera pointed
down towards the table.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{camera-placement.jpg}
	\caption{The camera angled downwards to capture the keyboard}
	\label{fig:metho-setup-camera-placement}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{actual-keeb.png}
	\caption{Image captured by the camera in one of the training iterations}
	\label{fig:metho-setup-keeb}
\end{figure}

The specific camera used was a Logitech C920. It is a 3 mega pixel webcam that
is capable of capturing color video in 1080p/30fps and 720p/30fps with a
diagonal field of view of 78\degree. This camera has a universal mounting clip
that allows the camera to be correctly positioned within the experimental setup
\parencite{logitech}.

@@

The keyboard used was a 60\% keyboard as shown in
Figure~\ref{fig:metho-setup-keeb}. This limited the necessary mapping for the
algorithm to the alphanumeric portion of the keyboard. This keyboard choice also
reduced the area that the camera captured, as this keyboard type is considerably
smaller compared to a full-size keyboard. The keyboard also had its keycaps and
case in a light color that contrasted the dark surface it was placed on. It also
had a dark USB-C cable to blend with the dark surface. These color coordination
steps improved initial keyboard detection. In addition, the keyboard layout was
ANSI, due to the availability and widespread adoption of the layout in the
Philippines.

\subsection{Environment}
The setup was lit with a lamp besides the camera. This light source evenly lit
the keyboard and the fingers used for typing. The light source used was a common
LED bulb rated at 9 Watts with 700 lumens. This light was white with a color
temperature of 6500k.

In addition, the surface where the keyboard was placed on was solid black
without any variation of color. This also improved initial keyboard detection.

@@

\subsection{Computer Vision based Keyboard Detection, and Key Mapping}
\label{section:metho-algo-keyboard}

A computer vision algorithm was created as a starting point
mapping the keys of the keyboard within a video. A rough flowchart of the algorithm
is shown in Figure~\ref{fig:metho-algo-key-flow}.

		\node[draw,
			below=of input,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (map) {Get Image Map};

		\node[draw,
			below=of map,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (contours) {Get Key Contour Points};

		\node[draw,
			trapezium,
			trapezium left angle = 65,
			trapezium right angle = 115,
			trapezium stretches,
			below=of contours,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (output) {Key-Edge Coordinatess Map};

		\draw[-latex] (start) edge (input)
		(input) edge (map)
		(map) edge (contours)
		(contours) edge (output)
		(output) edge (end);

\subsubsection{Get Image Map}
The first portion of the algorithm creates an image map that is overlaid over
the detected edges of the keyboard. The flowchart for this function is shown in
Figure~\ref{fig:metho-algo-key-image-map}.

@@

		\node[draw,
			below=of gray,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (denoise) {Denoising};

		\node[draw,
			below=of denoise,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (edge) {Edge Detection};

		\node[draw,
			below=of edge,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (thresh) {Thresholding};

		\node[draw,
			right=of gray,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (contours) {Find Contours};

		\node[draw,
			below=of contours,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (sort) {Simplify Contour};

		\node[draw,
			below=of sort,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (transform) {Transform Image Map};

		\node[draw,
			trapezium,
			trapezium left angle = 65,
			trapezium right angle = 115,
			trapezium stretches,
			below=of transform,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (output) {Virtual Image Map};

@@

		\node[coordinate, right = 0.5cm of thresh] (below) {};
		\node[coordinate, right = 0.5cm of gray] (top) {};

		\draw[-latex] (start) edge (input)
		(input) edge (gray)
		(gray) edge (denoise)
		(denoise) edge (edge)
		(edge) edge (thresh)
		(thresh) -- (below) |- (top) |- (contours)
		(contours) edge (sort)
		(sort) edge (transform)
		(transform) edge (output)
		(output) edge (end);

	\end{tikzpicture}
	\caption{Flowchart of Get Image Map}
	\label{fig:metho-algo-key-image-map}
\end{figure}

\paragraph{Convert to Grayscale}
The input frame from the camera was converted to a 256 level grayscale image
using \texttt{cvtColor} of OpenCV \parencite{opencv-cvtColor}. This step was
performed because future steps of the algorithm did not require color values to
work. In addition, this step optimized the algorithm as the number of dimensions
analyzed was reduced.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{grayscale.png}
	\caption{Camera capture converted to grayscale}
	\centering
\end{figure}

\paragraph{Denoising}
The algorithm denoised the grayscale image using a bilateral filter as
implemented by OpenCV \parencite{opencv-bilateral-filter}. This filter takes the
range of the image into account, rather than just the domain. This resulted in
an image that is smoothed while preserving its edges
\parencite{bilateral-filter}.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{denoised.png}
	\caption{Smoothed image with intact edges}
	\centering
\end{figure}

\paragraph{Edge Detection}
The algorithm used a Sobel filter for edge detection. This filter uses a
3$\times$3 kernel that is convolved twice. Once horizontally, and another
vertically to produce a grayscale image of the outlines within the frame. The
kernels used by the Sobel filter \parencite{sobel2014} is shown in
Figure~\ref{fig:metho-algo-key-sobel}.

\begin{figure}[H]
	\centering
	$\begin{bmatrix}
			+1 & 0 & -1 \\
			+2 & 0 & -2 \\
			+1 & 0 & -1
		\end{bmatrix}$
	$\begin{bmatrix}
			+1 & +2 & +1 \\
			0  & 0  & 0  \\
			-1 & -2 & -1
		\end{bmatrix}$
	\caption{Sobel Operator Kernels. Reproduced from \fullcite{sobel2014}}
	\label{fig:metho-algo-key-sobel}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{sobel.png}
	\caption{Output of the Sobel filter}
	\centering
\end{figure}

\paragraph{Thresholding}
The output of the Sobel filter is a grayscale image of 255 values, with each
gray pixel indicating edges within the image. There is a need to reduce the
range of these values to improve the ability of the algorithm to find the
contours of these edges. To do so, the algorithm utilized Otsu's algorithm to
perform automated thresholding. Otsu's algorithm determines a single threshold
that is most optimal for the image \parencite{otsu}. This outputs a
black-and-white image, with only two values.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{threshed.png}
	\caption{Threshed image of the edges}
	\centering
\end{figure}

\paragraph{Find Contours}
\label{section:metho-algo-key-contours}
Contours are curves joining all continuous points that have the same color or
intensity \parencite{opencv-contours}. The OpenCV function
\texttt{findContours}, with the \texttt{CHAIN\_APPROX\_SIMPLE} contour
approximation method, was used to find the contours of the outlines of the
object found using the previous step. This approximation method removed
redundant points and returned the least amount of points that describes the
shape. This OpenCV function implements the algorithm of~\cite{contours} in their
paper \citetitle{contours}.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{find-contours.png}
	\caption{All the contours found in the image}
	\centering
\end{figure}

\paragraph{Simplify Contour}
However, the points returned by the function can contain more than the four
extreme points at the edges of the keyboard. In addition, more than one contour
may be found within the image. As such, the algorithm sorted the contours by
area, and selected the largest one.

After sorting, the algorithm performed the Douglas-Peucker algorithm for Line
Simplification \parencite{douglas-peucker}. This reduced the number of points in
the largest contour to the minimum amount. In ideal cases, the number of points
would be four, with each point corresponding to the edges of the keyboard.
However, there were times when other objects would be within the frame, or they
intersected with the keyboard. This would result in a contour that would be
defined by more than four points --- which caused the entire algorithm to fail.

The contour points that described the contour were then sorted clockwise. This
was a necessity since the algorithm of~\cite{contours} does not guarantee that
the largest contour's points were returned in a clockwise arrangement, which is
a requirement for the next step

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{get-largest-contour.png}
	\caption{Simplified contour}
	\centering
\end{figure}

\paragraph{Transform Image Map}
The virtual keyboard map is a rectangular image that contains individual
\ac{roi} for each key in a 60\% \ac{ansi} keyboard. Each key has a corresponding
color assigned to it and this color fills the region where this key is located
at. Appendix Chapter~\ref{appendix:key-map} is a table that shows the hex color
code and key mapping present in the image map.

@@

The virtual map was stretched over the keyboard using OpenCV's
\texttt{warpPerspective} function. The function requires a perspective transform
that was calculated using \texttt{getPerspectiveTransform}. This function takes
in two arrays with four points each. The input points are the edges of the
virtual map, ordered clockwise. The output points are the four points of the
contour, ordered clockwise \parencite{opencv-image-transform}. The resulting
transform was then used by \texttt{warpPerspective} to apply the transform to
the virtual map.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.6\textwidth]{transformed-image-map.png}
	\caption{Transformed image map}
	\centering
\end{figure}

\subsubsection{Get Key Contour Points}
The second portion of the algorithm pre-computes the contour points of each key
in the keyboard based on the generated virtual map and the Key-Color Values map.
The flowchart of this function is shown in
Figure~\ref{fig:metho-algo-key-contour-points}

		\node[draw,
			trapezium,
			trapezium left angle = 65,
			trapezium right angle = 115,
			trapezium stretches,
			below = of start,
			minimum width = 3.5cm,
			minimum height = 1cm,
			align = center,
			inner sep = 3mm
		] (input) {Virtual Map \\ Key-Color Values Map};

		\node[draw,
			diamond,
			aspect = 2,
			below = of input,
			minimum width = 4cm,
			inner sep = 2mm] (for) {for Key, Color in Key-Color Values Map};

		\node[draw,
			below=of for,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (filter) {Filter Virtual Map};

		\node[draw,
			below=of filter,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (contours) {Find and Simplify Contour};

		\node[draw,
			below=of contours,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (scale) {Scale Contours};

		\node[draw,
			trapezium,
			trapezium left angle = 65,
			trapezium right angle = 115,
			trapezium stretches,
			below=of scale,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (output) {Key-Edge Coordinatess Map};

		\node[coordinate, right = 1cm of for] (top-right) {};
		\node[coordinate, right = 1cm of output] (below-right) {};
		\node[coordinate, left = 1cm of for] (top-left) {};
		\node[coordinate, left = 2cm of scale] (below-left) {};

		\draw[-latex] (start) edge (input)
		(input) edge (for)
		(for) edge (filter)
		(for) -- (top-right) -- (below-right) |- (output)
		(filter) edge (contours)
		(contours) edge (scale)
		(scale) -- (below-left) -- (top-left) |- (for)
		(output) edge (end);

	\end{tikzpicture}
	\caption{Flowchart of Get Key Contour Points}
	\label{fig:metho-algo-key-contour-points}
\end{figure}

\paragraph{Filter Virtual Map}
The color that corresponds to each key in the virtual map is stored in the
Key-Color Value Map. This color was used to create a mask for the virtual map
for a specific key. This mask was obtained by using this snippet of code
\texttt{(virtualMap == key).all(axis=2)}. This mask was then used to black out
the rest of the virtual map, leaving only the \ac{roi} of the key in the image.

This specific method was used as the other method that was trialed,
\texttt{np.where(virtualMap != key)}, did not consider all 3 channels when
comparing each pixel --- i.e.\ a key assigned with the color value of
\texttt{[100, 0, 0]} will not be blacked out if the color value of the key to be
isolated is \texttt{[100, 243, 0]} as the blue channel of the pixel,
\texttt{100} is equal in both. This would result in other keys remaining in the
virtual map, even if only one key corresponds to that color.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{filtered-key.png}
	\caption{Filtered key}
	\centering
\end{figure}

\paragraph{Find and Simplify Contours}
The same method in finding and simplifying the contour of the keyboard was used
to find and simplify the contour of the \ac{roi} of the key.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{key-contour.png}
	\caption{Contour points of the \ac{roi}}
	\centering
\end{figure}

\paragraph{Scale Contours}
During testing using training data, a lot of the failures in finger-key
identification was due to the tight fit of the contour to the key. The contour
on its own did not give enough buffer for the placement of the finger. This
buffer was needed as, in some instances, the finger used to press the key was
not exactly on top of the key, but rather to its side. The algorithm accounts
for this situation by adding a buffer. This was achieved through scaling the
contour points by seven pixels on each side. This value was obtained after
testing other possible values. A buffer of five pixels did not have a lot of
effect in reducing the number of failures. A buffer of ten pixels did reduce the
number of failures, but it also greatly increased the algorithm's uncertainty in
determining the finger --- uncertainty is defined as detecting two or more
fingers within one \ac{roi}. Seven was a good middle ground in decreasing
failures, without greatly increasing uncertainty. More information about this
can be found in the Results and Discussion at Section~\ref{section:results}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{scaled-contour.png}
	\caption{Contour points of the \ac{roi}}
	\centering
\end{figure}

\paragraph{Key-Edge Coordinatess Map}
After all the keys have passed through the loop, a Key-Edge Coordinates Map was
generated. Each key now has four points that serve as the coordinates of the
edge of its \ac{roi}.

The coordinate system that the contour points are based on had its origin at the
top left, starting at 0, 0. Going to the right increased the value of the
x-axis, and going to the bottom increased the value of the y-axis. This results
in a coordinate system that only has positive values with each value
corresponding to a pixel.

\subsection{Computer Vision based Finger Detection, and Tracking}
\label{section:metho-algo-finger}
MediaPipe Hands was used as the finger detection and tracking solution. This
algorithm is composed of two ML models working in conjunction to be able to
detect the different parts of the hands and track them accurately.

@@

The two previously chosen algorithms was combined to accomplish finger-key
identification and mapping.

The integration of the algorithm was a two-step process. The first step was to
get the Key-Edge Coordinates Map shown in
Section~\ref{section:metho-algo-keyboard}.

The second step runs whenever a key press has been detected. This step used the
Key-Edge Coordinates Map in conjunction with the finger tracking algorithm selected
in Section~\ref{section:metho-algo-finger}. The flowchart of the algorithm is
shown in Figure~\ref{fig:metho-algo-integration}

@@

		\node[draw,
			trapezium,
			trapezium left angle = 65,
			trapezium right angle = 115,
			trapezium stretches,
			below = 2cm of if-packet,
			minimum width = 3.5cm,
			minimum height = 1cm,
			align = center,
			inner sep = 3mm
		] (map) {Key-Edge Coordinates Map \\ Finger Tracking Data};

		\node[draw,
			below=of map,
			minimum width = 3.5cm,
			minimum height = 1cm,
			inner sep = 3mm
		] (find-finger) {Find which fingertip is on that \ac{roi}};

@@

		\draw[-latex] (start) edge (if-packet)
		(map) edge (find-finger)
		(find-finger) edge (return)
		(return) edge (end);

@@

\subsubsection{Find which fingertip is on that \ac{roi}}
The Finger Tracking Data contained the pixel positions of each landmark of the
hand. For this step, the algorithm finds the landmark which is positioned within
the single \ac{roi} obtained from the previous step. This was done using a
series of checks.

Each landmark's coordinates was compared to the coordinates of the edges of the
\ac{roi}. A landmark was determined as positioned within the \ac{roi} if all the
following conditions are true:
\begin{enumerate}
	\item In the X axis, the landmark's coordinates is greater than one or both of
	      the two coordinates found of the left side of the \ac{roi}
	\item In the X axis, the landmark's coordinates is less than one or both of
	      the two coordinates found of the right side of the \ac{roi}
	\item In the Y axis, the landmark's coordinates is greater than one or both of
	      the two coordinates found of the top side of the \ac{roi}
	\item In the Y axis, the landmark's coordinates is less than one or both of
	      the two coordinates found of the bottom side of the \ac{roi}
\end{enumerate}

These conditions maximized the total area of the \ac{roi}, and it is not strict
about exact accuracy. In essence, these conditions creates a perfectly
rectangular box that contained the quadrilateral formed by the four points from
the coordinates.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{detected-finger.png}
	\caption{Finger tip within an \ac{roi}}
	\centering
\end{figure}

\subsubsection{Return which fingertip}
The landmark was then used to determine which specific finger corresponds to the
key press. Figure~\ref{fig:metho-algo-integration-landmarks} shows each possible
landmark that may be returned from the previous step. This step returned the
name of the landmark, up until the first underscore. As an example, if the
landmark found is \texttt{MIDDLE\_FINGER\_TIP}, this step will return
\texttt{MIDDLE} denoting that the middle finger is the finger that corresponds
with the key press. In addition, the specific hand will also be returned as one
of two strings, \texttt{LEFT} and \texttt{RIGHT}, since this information is also
bundled together with the landmarks. The two strings were then concatenated with
an underscore. An example return value is \texttt{RIGHT\_RING}.

@@

\subsection{Implementation}
\label{section:metho-algo-implementation}
The previously discussed algorithms was implemented using Python. The Keyboard
Detection, and Mapping Algorithm in Section~\ref{section:metho-algo-keyboard}
was implemented as a separate submodule in Python using OpenCV. The Finger
Detection, and Tracking Solution in Section~\ref{section:metho-algo-finger} was
consumed using the prebuilt Python package offered by the MediaPipe team.
Finally, the integration of the two will be done as another separate submodule
in Python.

\section{Training and Testing}

\subsection{Typing Test Sequences}
Typing test sequences are strings that was used in testing the user in their
ability to type. The test sequences that was used came from text found in the
public domain obtained from Project Gutenberg and the Internet Archive.
Sentences were isolated from these text and used as test sequences if they fit
the criteria.

The criteria for choosing test sequences were as follows: (1) $\approx80\%$ of the
characters in the keyboard is present in a test sequence. (2) The number of
words in a test sequence do not exceed 25. (3) Numbers and punctuations should
be present in at least $\approx30\%$ of the total test sequences.

There was a total of 10 test sequences that was gathered. An example test
sequence is ``What of it, if some old hunks of a sea-captain orders me to get a
broom and sweep down the decks?'' from Moby Dick by~\cite{moby-dick}. The 9
other test sequences can be found in Appendix
Chapter~\ref{appendix:test-sequences}.

\subsection{Data Gathering}
\subsubsection{Video Capture}
There were 3 groups of 10 videos that were captured by the researcher, for a
total of 30 videos. Each group corresponds to three predetermined speeds of
typing: slow (15wpm), average (35wpm), and fast (80wpm). These speeds were based
on test data from~\cite{keybr}. The researcher typed the 10 typing test
sequences at these predetermined speeds. During typing, errors were not
consciously taken into account, and errors happened naturally as part of the
typing process.

A Python script was created to facilitate this process. The script captured a
video of the researcher as the researcher was typing the test sequences.
Whenever the researcher pressed a key, the corresponding frame count was
captured, and the key pressed was recorded in conjunction. This was stored in
the format \texttt{FRAME\_NO:KEY\_PRESSED}, and each keypress was then stored as
a line in a file.

\subsubsection{Keypress Labelling}
The researcher then manually perform finger-key identification for all key
presses present in the video. This was also done through a Python script. The
script parsed the file associated with a video. The frame number found at each
line was then shown, and the associated data with the frame was superimposed
over it. The researcher then pressed keys that corresponded with the finger used
to press the key. This was then again stored in the format
\texttt{FRAME\_NO:KEY\_PRESSED:FINGER\_USED}, and each keypress was then stored
as a line in a file.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.6\textwidth]{labelling.png}
	\caption{Screen showing the labelling process}
	\centering
\end{figure}

\subsubsection{Test-Training Split}
The annotated data was then divided into training and test data at a ratio of
60:40.
